Raft Distributed Consensus Algorithm over\\ WebRTC (Web Real-Time Communication)

### 2015-12-07 research_class/paper-min.tex
Add workshop.tex, paper-min.tex and update data files.

\begin{enumerate}
\item The paper describes the Raft RPCs in terms of synchronous stream
    based request/response model where requests and responses are
    automatically correlated together by a lower level of the software
    stack. For example with HTTP 1.0 \cite{http:rfc1945} client
    requests are followed by a server response on any given TCP
    stream. Most remote procedure call mechanims also use
    a synchronous request/response model. %TODO/CITE
    WebRTC DataChannel connections are message based rather than
    request/response stream based. %TODO/CITE
    In the original Raft definition of RPC, request and response messages
    almost contain enough information to function correctly in an
    environment without builtin request/response correlation, such as
    UDP or SCTP when used with reliable transport disabled. %TODO/CITE
    Raft.js includes a small modification to the response RPCs for
    requestVote and appendEntries to add a `sourceId' field that
    enables the protocol to function with message based transports.
\item The Raft paper is clear that each membership change must be
    delayed until all the previous membership change is committed
    (sequential) \cite[Section~4.1]{raft_thesis:ongaro14}. It is
    implied that the leader node should track all pending membership
    change requests and apply them one at a time.  However, in
    Raft.js, if a client requests a membership change while another
    change request is pending (uncommitted), the leader will reject
    the change request with a status of `PENDING\_CONFIG\_CHANGE'.
    This keeps the core Raft implementation simpler with the trade-off
    being that the client implementation of membership change requests
    is slightly more complex due to retry logic.
\end{enumerate}

@@

@@

To simplify the execution of the tests and to make the testing
infrastructure more portable, a docker container environment was used
to encapsulate and execute the SlimerJS environment. The container is
built using \emph{test/Dockerfile} which is in turn derived from
the ``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
This custom Dockerfile was created because the upstream version of the
Dockerfile used an older version of XULRunner that did not support
more than 7 WebRTC clients communicating together simultaneously. This
issue was discovered during testing.

Writing tests using SlimerJS involves creating JavaScript programs
that are run within the SlimerJS environment. These scripts can create
browser instances (essentially headless browser tabs), query the status
of those pages and execute JavaScript within the context of the pages
themselves.

Three different tests were created to test the basic Raft over WebRTC
application and the chat application. Each of the tests takes
a parameter indicating the number of nodes/pages to launch and test
again. The test script files are:

\begin{enumerate}
    \item \emph{test/common.js}: a module of common functions used by
        the other tests scripts.
    \item \emph{test/test\_up.js}: launch and start an initial leader
        node/page, then start the rest of the nodes (if any) so that
        the leader will configure them into the cluster. Then test
        then waits for every node to be recorded (applied) into the log
        of every other node of the cluster. The test reports the
        number of milliseconds for the cluster to reach this state.
    \item \emph{test/test\_kill\_nodes.js}: similar to
        \emph{test/test\_up.js} but once all the nodes are configured
        into the cluster, this script then kills the leader plus
        a configurable number of nodes. By default the script kills
        just under half the nodes of the cluster. The test then waits
        for a new leader to be elected and remove the dead nodes from
        the cluster. The test reports both the number of milliseconds
        for the cluster to come up and the recover from the failure.
    \item \emph{test/test\_chat\_propagate.js}: similar to
        \emph{test/test\_up.js} but once the cluster is up, the test
        connects to one of the non-leader nodes and and sends a chat
        text line. The test then waits for the chat message to
        propagate to every other node's state machine. The test
        reports both the number of milliseconds for the cluster to
        come up and for the messages to propagate to every node.
\end{enumerate}

@@

Three tests were run at 10 different cluster sizes (odd cluster sizes
from 3 to 21 nodes): \emph{test/test\_up.js} and
\emph{test/test\_kill\_nodes.js} in both modes (kill only leader and kill just
under half the nodes).
Each of these tests was run with three different Raft election
timeout values: 100ms, 500ms, 1000ms. The
\emph{test/test\_chat\_propagate.js} test was run at
11 different cluster sizes (odd cluster sizes from 1 to 21 nodes).
Each test mode and cluster size was run 7 times for a total of 707
test runs (3*10*7*3 + 11*7)

@@

Figure~\ref{fig:propagate1} shows the results of running the
\emph{test/wait\_chat\_propagate.js} test to show the amount of time
that a stable cluster takes to propagate, commit and apply a single
chat message to the state machine on every cluster node using the
atomic \emph{seqAppend} command.  Figure~\ref{fig:propagate1} is
a scatterplot with each test run as a blue ``x" and a black line
representing the average at each cluster size.

\begin{figure}[!t]
    \begin{tikzpicture}
    \begin{axis}[%
        xlabel={Nodes}, ylabel={Time (seconds)},
        %xtick=data,
        xmin=0, ymin=0,
        grid=major,
        scatter/classes={%
            propagate_1_et1000={mark=x,draw=blue},
            propagate_1_et1000_average={mark=}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_1.dat};
    \addplot[scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_1_average.dat};
    \end{axis}
    \end{tikzpicture}
\caption{Propagate$,$ Commit$,$ and Apply Command}
  \label{fig:propagate1}
\end{figure}

Figure~\ref{fig:propagate100} is similar to
Figure~\ref{fig:propagate1} but shows the results of adding 100 chat
messages instead of just 1. The commands are serialized which means
that the previous command must be committed before the current command
can be committed. The test is complete once the last chat message
command has been both committed and applied to every node of the
cluster. Figure~\ref{fig:propagate1} is a scatterplot with each test
run as a red ``x" and a black line representing the average at each
cluster size.

\begin{figure}[!t]
    \begin{tikzpicture}
    \begin{axis}[%
        xlabel={Nodes}, ylabel={Time (seconds)},
        %xtick=data,
        xmin=0, ymin=0,
        grid=major,
        scatter/classes={%
            propagate_100_et1000={mark=x,draw=red},
            propagate_100_et1000_average={mark=}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_100.dat};
    \addplot[scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_100_average.dat};
    \end{axis}
    \end{tikzpicture}
\caption{Propagate$,$ Commit$,$ and Apply Command}
  \label{fig:propagate100}
\end{figure}

Figures \ref{fig:cluster_up}, \ref{fig:recovery-leader-failure} and
\ref{fig:recovery-half-failure} show the results of running the
\emph{test/test\_up.js} and \emph{test/test\_kill\_nodes.js} in both
modes (kill only leader and kill just under half the nodes). Three
different scatterplots are shown for each test mode to test different
values of the Raft election timeout value: 100ms, 500ms, and 1000ms.
The election timeout value makes very little difference to the average
time curve line. The reason for this is that the first leader tends to
stay the leader as new nodes are added so for the most part the
election timer does not have an impact.

Some specific notes about the figures:

% TODO: link figure numbers

\begin{enumerate}
    \item \emph{Figure~\ref{fig:cluster_up}~(a)}: The results show that
        once the cluster size reaches 17, the cluster sometimes does
        not activate before the test times out (the red ``x" marks).
        Once the cluster reaches 21 nodes, all 7 test runs time out.
        The reason for this is because at 17 nodes, the leader is not
        always able to send out appendEntries RPCs to all the other
        nodes before their election timeout fires. More analysis is
        needed, but one possible explanation is that the single
        SlimerJS XULRunner process that is handling all the cluster
        nodes is not able to give enough CPU time to the leader node
        to reliably send appendEntries RPCs to all the other nodes
        within the election timeout value. Increasing the election
        timeout to 500ms or greater eliminates the failures without
        significantly changing the average cluster activation time
        curve.
    \item \emph{Figure~\ref{fig:recovery-leader-failure}~(a)}: The
        results show that when the cluster size reaches 15, then
        average amount of time for cluster recovery (after the leader
        is killed) increases dramatically (not all data points are
        shown on the plot for size 15). At cluster size 17 the cluster
        is no longer able to recover within the test timeout period.
        Increasing the election timeout to 500ms or greater eliminates
        the failures at the cost of slightly increased recovery time
        for smaller cluster sizes.
    \item \emph{Figure~\ref{fig:recovery-half-failure}~(a)}: During
        this test the cluster was unable to activate at cluster size
        21 when the election timeout value is set to 100ms (see
        Figure~\ref{fig:cluster_up}~(a)). This means there was no
        opportunity for cluster recovery at cluster size 21 so there
        is no data for that column.

Each figure is a scatterplot with each test run as a blue ``x" and
a black line representing the average at each cluster size.

%\begin{figure*}[!t]
\begin{figure*}[h]
    \centering
    %\begin{subfigure}[b]{0.30\textwidth}
    \subfloat[100ms election timer]{
        \centering
        \resizebox{0.3\linewidth}{!}{
            \begin{tikzpicture}
            \begin{axis}[%
                %height=1.5\textwidth, width=1.4\textwidth,
                xlabel={Nodes}, ylabel={Time (seconds)},
                xtick=data,
                ymin=0, ymax=65,
                grid=major,
                scatter/classes={%
                    et100={mark=x,draw=blue},
                    et100_fail={mark=x,draw=red},
                    et100_average={mark=,draw=red}}]
            \addplot[scatter,only marks,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_cluster_up_100.dat};
            \addplot[scatter,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_cluster_up_100_averages.dat};
            \end{axis}
            \end{tikzpicture}
        }
        \label{fig:A}
    }
    %\begin{subfigure}[b]{0.30\textwidth}
    \subfloat[500ms election timer]{
        \centering
        \resizebox{0.3\linewidth}{!}{
            \begin{tikzpicture}
            \begin{axis}[%
                %height=1.5\textwidth, width=1.4\textwidth,
                xlabel={Nodes}, ylabel={Time (seconds)},
                xtick=data,
                ymin=0, ymax=65,
                grid=major,
                scatter/classes={%
                    et500={mark=x,draw=blue},
                    et500_average={mark=,draw=red}}]
            \addplot[scatter,only marks,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_cluster_up_500.dat};
            \addplot[scatter,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_cluster_up_500_averages.dat};
            \end{axis}
            \end{tikzpicture}
        }
        \label{fig:B}
    }
    %\begin{subfigure}[b]{0.30\textwidth}
    \subfloat[1000ms election timer]{
        \centering
        \resizebox{0.3\linewidth}{!}{
            \begin{tikzpicture}
            \begin{axis}[%
                %height=1.5\textwidth, width=1.4\textwidth,
                xlabel={Nodes}, ylabel={Time (seconds)},
                %xtick={1,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33},
                %xticklabels={1,,5,,9,,13,,17,,21,,25,,29,,33},
                xmin=0, xmax=34,
                ymin=0, ymax=11,
                grid=major,
                scatter/classes={%
                    start_et1000={mark=x,draw=blue},
                    start_et1000_average={mark=,draw=red}}]
            \addplot[scatter,only marks,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/start_et1000.dat};
            \addplot[scatter,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/start_et1000_average.dat};
            \end{axis}
            \end{tikzpicture}
        }
        \label{fig:C}
    }
    \caption{Initial Cluster Activation}
    \label{fig:cluster_up}

%%% Next Row
\rule{\textwidth}{0.3pt}
\smallskip

    %\begin{subfigure}[b]{0.30\textwidth}
    \subfloat[100ms election timer]{
        \centering
        \resizebox{0.3\linewidth}{!}{
            \begin{tikzpicture}
            \begin{axis}[%
                %height=1.5\textwidth, width=1.4\textwidth,
                xlabel={Nodes}, ylabel={Time (seconds)},
                xtick=data,
                ymin=0, ymax=3.5, xmax=23,
                grid=major,
                scatter/classes={%
                    kl={mark=x,draw=blue},
                    kl_fail={mark=x,draw=red},
                    kl_average={mark=,draw=red}}]
            \addplot[scatter,only marks,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_kill_leader_100.dat};
            \addplot[scatter,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_kill_leader_100_averages.dat};
            \end{axis}
            \end{tikzpicture}
        }
        \label{fig:D}
    }
    %\begin{subfigure}[b]{0.30\textwidth}
    \subfloat[500ms election timer]{
        \centering
        \resizebox{0.3\linewidth}{!}{
            \begin{tikzpicture}
            \begin{axis}[%
                %height=1.5\textwidth, width=1.4\textwidth,
                xlabel={Nodes}, ylabel={Time (seconds)},
                xtick=data,
                ymin=0, ymax=3.5,
                grid=major,
                scatter/classes={%
                    kl={mark=x,draw=blue},
                    kl_average={mark=,draw=red}}]
            \addplot[scatter,only marks,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_kill_leader_500.dat};
            \addplot[scatter,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_kill_leader_500_averages.dat};
            \end{axis}
            \end{tikzpicture}
        }
        \label{fig:E}
    }
    %\begin{subfigure}[b]{0.30\textwidth}
    \subfloat[1000ms election timer]{
        \centering
        \resizebox{0.3\linewidth}{!}{
            \begin{tikzpicture}
            \begin{axis}[%
                %height=1.5\textwidth, width=1.4\textwidth,
                xlabel={Nodes}, ylabel={Time (seconds)},
                %xtick=data,
                xmin=0,
                ymin=0, ymax=3.5,
                grid=major,
                scatter/classes={%
                    kill_1_et1000={mark=x,draw=blue},
                    kill_1_et1000_average={mark=,draw=red}}]
            \addplot[scatter,only marks,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/kill_1_et1000.dat};
            \addplot[scatter,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/kill_1_et1000_average.dat};
            \end{axis}
            \end{tikzpicture}
        }
        \label{fig:F}
    }
    \caption{Recovery from Leader Failure}
    \label{fig:recovery-leader-failure}

%%% Next Row
\rule{\textwidth}{0.3pt}
\smallskip

    %\begin{subfigure}[b]{0.30\textwidth}
    \subfloat[100ms election timer]{
        \centering
        \resizebox{0.3\linewidth}{!}{
            \begin{tikzpicture}
            \begin{axis}[%
                %height=1.5\textwidth, width=1.4\textwidth,
                xlabel={Nodes}, ylabel={Time (seconds)},
                xtick=data,
                ymin=0, ymax=3.5, xmax=23,
                grid=major,
                scatter/classes={%
                    kh={mark=x,draw=blue},
                    kh_average={mark=,draw=red}}]
            \addplot[scatter,only marks,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_kill_half_100.dat};
            \addplot[scatter,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_kill_half_100_averages.dat};
            \end{axis}
            \end{tikzpicture}
        }
        \label{fig:G}
    }
    %\begin{subfigure}[b]{0.30\textwidth}
    \subfloat[500ms election timer]{
        \centering
        \resizebox{0.3\linewidth}{!}{
            \begin{tikzpicture}
            \begin{axis}[%
                %height=1.5\textwidth, width=1.4\textwidth,
                xlabel={Nodes}, ylabel={Time (seconds)},
                xtick=data,
                ymin=0, ymax=3.5,
                grid=major,
                scatter/classes={%
                    kh={mark=x,draw=blue},
                    kh_average={mark=,draw=red}}]
            \addplot[scatter,only marks,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_kill_half_500.dat};
            \addplot[scatter,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/data_kill_half_500_averages.dat};
            \end{axis}
            \end{tikzpicture}
        }
        \label{fig:H}
    }
    %\begin{subfigure}[b]{0.30\textwidth}
    \subfloat[1000ms election timer]{
        \centering
        \resizebox{0.3\linewidth}{!}{
            \begin{tikzpicture}
            \begin{axis}[%
                %height=1.5\textwidth, width=1.4\textwidth,
                xlabel={Nodes}, ylabel={Time (seconds)},
                xmin=0,
                ymin=0, ymax=3.5,
                grid=major,
                scatter/classes={%
                    kill_half_et1000={mark=x,draw=blue},
                    kill_half_et1000_average={mark=,draw=red}}]
            \addplot[scatter,only marks,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/kill_half_et1000.dat};
            \addplot[scatter,%
                scatter src=explicit symbolic]%
            table[meta=label] {data/kill_half_et1000_average.dat};
            \end{axis}
            \end{tikzpicture}
        }
        \label{fig:I}
    }
    \caption{Recovery from Half Cluster Failure}
    \label{fig:recovery-half-failure}
\end{figure*}

>>>>

\begin{enumerate}
\item The paper describes the Raft RPCs in terms of synchronous stream
    based request/response model where requests and responses are
    automatically correlated together by a lower level of the software
    stack. For example with HTTP 1.0 \cite{http:rfc1945} client
    requests are followed by a server response on any given TCP
    stream. Most remote procedure call mechanisms also use
    a synchronous request/response model. %TODO/CITE
    WebRTC DataChannel connections are message based rather than
    request/response stream based. %TODO/CITE
    In the original Raft definition of RPC, request and response messages
    almost contain enough information to function correctly in an
    environment without builtin request/response correlation, such as
    UDP or SCTP when used with reliable transport disabled. %TODO/CITE
    Raft.js includes a small modification to the response RPCs for
    requestVote and appendEntries to add a `sourceId' field that
    enables the protocol to function with message based transports.
\item The Raft paper is clear that each membership change must be
    delayed until all the previous membership change is committed
    (sequential) \cite[Section~4.1]{raft_thesis:ongaro14}. It is
    implied that the leader node should track all pending membership
    change requests and apply them one at a time.  However, in
    Raft.js, if a client requests a membership change while another
    change request is pending (uncommitted), the leader will reject
    the change request with a status of `PENDING\_CONFIG\_CHANGE'.
    This keeps the core Raft implementation simpler with the trade-off
    being that the client implementation of membership change requests
    is slightly more complex due to retry logic.
\end{enumerate}

@@

\subsubsection{Twst Test Framework}

In order to facilitate coordinated testing across a large number of
browser instances, a new test framework was created called
\cite{twst}. This \emph{Twst} framework provides a client and server
library written in JavaScript:

\begin{enumerate}
    \item \emph{server library}: this library is used to write tests
        that need to coordinate testing across many browser instances.
        It provides services such as client tracking/enumeration,
        broadcast and targeted RPC, state rendezvous, log collection,
        etc.
    \item \emph{client library}: this library is included with the web
        application that is being tested. When the page loads, the
        library will automatically connect to the \emph{Twst} server
        instance and begin sending console logs and accepting
        RPCs/commands from the server.
\end{enumerate}

\subsubsection{Test Setup}

@@

To simplify the execution of the tests and to make the testing
infrastructure more portable, a docker container environment was used
to encapsulate and execute the SlimerJS environment. The container is
built using \emph{test/Dockerfile} which is in turn derived from
the ``fentas/slimerjs" docker image \cite{fentas:slimerjs}.
This custom Dockerfile was created because the upstream version of the
Dockerfile used an older version of XULRunner that did not support
more than 7 WebRTC clients communicating together simultaneously. This
issue was discovered during initial testing.

Writing tests using SlimerJS involves creating JavaScript test programs
which start a \emph{Twst} server instance and then launch a SlimerJS
docker container for each node of the cluster. Each container starts
a XULRunner browser instance of the Raft over WebRTC chat application.
Also included with the application is the \emph{Twst} client library
which connects back to the \emph{Twst} service in the test program.
Once all the chat application nodes have connected to the test
server/program, then the actual test is started across all the
application nodes.

Three different tests programs/servers were created to test the basic
Raft over WebRTC application. Each of the tests takes a parameter
indicating the number of nodes/pages to launch. The test
source files are:

\begin{enumerate}
    \item \emph{test/common.js}: a module of common functions used by
        the other tests scripts.
    \item \emph{test/test\_up.js}: launch and start an initial leader
        node/page, then start the rest of the nodes (if any) so that
        the leader will configure them into the cluster. Then test
        then waits for every node to be recorded (applied) into the log
        of every other node of the cluster. The test reports the
        number of milliseconds for the cluster to reach this state.
    \item \emph{test/test\_kill\_nodes.js}: similar to
        \emph{test/test\_up.js} but once all the nodes are configured
        into the cluster, this script then kills the leader plus
        a configurable number of nodes. By default the script kills
        just under half the nodes of the cluster. The test then waits
        for a new leader to be elected and remove the dead nodes from
        the cluster. The test reports both the number of milliseconds
        for the cluster to come up and the number of milliseconds for
        the cluster to fully recover from the failure and evict the
        failed nodes.
    \item \emph{test/test\_chat\_propagate.js}: similar to
        \emph{test/test\_up.js} but once the cluster is up, the test
        connects to one of the non-leader nodes and and sends a chat
        text line. The test then waits for the chat message to
        propagate to every other node's state machine. The test
        reports both the number of milliseconds for the cluster to
        come up and for the messages to propagate to every node.
\end{enumerate}

@@

Five different tests (using different options for the three test
programs) were run at 17 different cluster sizes (odd cluster sizes
from 3 to 35 nodes). Each of these tests was run with three different
Raft election timeout values: 250ms, 500ms, 1000ms. Each of the five
tests and cluster size combinations was run 7 times.

@@

\begin{figure}[!t]
    \begin{tikzpicture}
    \begin{axis}[%
        xlabel={Nodes}, ylabel={Time (seconds)},
        %xtick=data,
        xmin=0, ymin=0,
        grid=major,
        scatter/classes={%
            propagate_1_et1000={mark=x,draw=blue},
            propagate_1_et1000_average={mark=}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_1.dat};
    \addplot[scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_1_average.dat};
    \end{axis}
    \end{tikzpicture}
    \caption{Propagate$,$ Commit$,$ and Apply 1 Command}
    \label{fig:propagate1}
%
    \vspace{1cm}
%
    \begin{tikzpicture}
    \begin{axis}[%
        xlabel={Nodes}, ylabel={Time (seconds)},
        %xtick=data,
        xmin=0, ymin=0,
        grid=major,
        scatter/classes={%
            propagate_100_et1000={mark=x,draw=blue},
            propagate_100_et1000_average={mark=}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_100.dat};
    \addplot[scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/propagate_100_average.dat};
    \end{axis}
    \end{tikzpicture}
    \caption{Propagate$,$ Commit$,$ and Apply 100 Commands}
    \label{fig:propagate100}
\end{figure}

\begin{figure}[!t]
    \begin{tikzpicture}
    \begin{axis}[%
        %height=1.5\textwidth, width=1.4\textwidth,
        %width=0.45\textwidth,
        xlabel={Nodes}, ylabel={Time (seconds)},
        xmin=0, xmax=43,
        ymin=0, ymax=15,
        grid=major,
        %legend pos=outer north east,
        legend pos=north west,
        scatter/classes={%
            start_et250={mark=x,draw=magenta},
            start_et250_average={mark=,draw=red},
            %start_et250_fail={mark=x,/raw=red},
            start_et500={mark=x,draw=cyan},
            start_et500_average={mark=,draw=red},
            %start_et500_fail={mark=x,draw=red},
            start_et1000={mark=x,draw=blue},
            start_et1000_average={mark=,draw=red}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/start_et250.dat};
    \addplot[magenta,scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/start_et250_average.dat};

    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/start_et500.dat};
    \addplot[cyan,scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/start_et500_average.dat};
    %\addplot[scatter,%
    %    scatter src=explicit symbolic]%
    %table[meta=label] {data/start_et500_fail.dat};

    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/start_et1000.dat};
    \addplot[blue,scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/start_et1000_average.dat};

    \legend{250ms,,500ms,,1000ms,}
    \end{axis}
    \end{tikzpicture}
    \caption{Initial Cluster Activation}
    \label{fig:cluster_up}
\end{figure}

\begin{figure}[!t]
    \begin{tikzpicture}
    \begin{axis}[%
        %height=1.5\textwidth, width=1.4\textwidth,
        %width=0.45\textwidth,
        xlabel={Nodes}, ylabel={Time (seconds)},
        xmin=0, xmax=37,
        ymin=0, ymax=4.5,
        grid=major,
        %legend pos=outer north east,
        legend pos=north west,
        scatter/classes={%
            kill_1_et250={mark=x,draw=magenta},
            kill_1_et250_average={mark=,draw=red},
            %kill_1_et250_fail={mark=x,/raw=red},
            kill_1_et500={mark=x,draw=cyan},
            kill_1_et500_average={mark=,draw=red},
            %kill_1_et500_fail={mark=x,draw=red},
            kill_1_et1000={mark=x,draw=blue},
            kill_1_et1000_average={mark=,draw=red}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_1_et250.dat};
    \addplot[magenta,scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_1_et250_average.dat};

    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_1_et500.dat};
    \addplot[cyan,scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_1_et500_average.dat};
    %\addplot[scatter,%
    %    scatter src=explicit symbolic]%
    %table[meta=label] {data/kill_1_et500_fail.dat};

    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_1_et1000.dat};
    \addplot[blue,scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_1_et1000_average.dat};

    \legend{250ms,,500ms,,1000ms,}

    \end{axis}
    \end{tikzpicture}
    \caption{Recovery from Leader Failure}
    \label{fig:recovery_leader_failure}
%
    \vspace{1cm}
%
    \begin{tikzpicture}
    \begin{axis}[%
        %height=1.5\textwidth, width=1.4\textwidth,
        %width=0.45\textwidth,
        xlabel={Nodes}, ylabel={Time (seconds)},
        xmin=0, xmax=37,
        ymin=0, ymax=4.5,
        grid=major,
        %legend pos=outer north east,
        legend pos=north west,
        scatter/classes={%
            kill_half_et250={mark=x,draw=magenta},
            kill_half_et250_average={mark=,draw=red},
            %kill_half_et250_fail={mark=x,/raw=red},
            kill_half_et500={mark=x,draw=cyan},
            kill_half_et500_average={mark=,draw=red},
            %kill_half_et500_fail={mark=x,draw=red},
            kill_half_et1000={mark=x,draw=blue},
            kill_half_et1000_average={mark=,draw=red}}]
    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_half_et250.dat};
    \addplot[magenta,scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_half_et250_average.dat};

    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_half_et500.dat};
    \addplot[cyan,scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_half_et500_average.dat};
    %\addplot[scatter,%
    %    scatter src=explicit symbolic]%
    %table[meta=label] {data/kill_half_et500_fail.dat};

    \addplot[scatter,only marks,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_half_et1000.dat};
    \addplot[blue,scatter,%
        scatter src=explicit symbolic]%
    table[meta=label] {data/kill_half_et1000_average.dat};

    \legend{250ms,,500ms,,1000ms,}

    \end{axis}
    \end{tikzpicture}
    \caption{Recovery from Half Cluster Failure}
    \label{fig:recovery_half_failure}
\end{figure}

Figure~\ref{fig:propagate1} and Figure~\ref{fig:propagate100} show the
result of running the \emph{test/wait\_chat\_propagate.js} test to
determine the amount of time that a stable cluster takes to propagate,
commit and apply 1 and 100 chat message commands respectively to the state
machine on every cluster node using the atomic \emph{seqAppend}
command. The commands are serialized which means that the previous
command must be committed before the current command can be committed.
The test is complete once the last chat message command has been both
committed and applied to every node of the cluster. Each test execution
is shown as a blue ``x" and the black line represents the average for
each cluster size.

In Figure~\ref{fig:propagate1} the time to propagate a single message
to a single node of the cluster is approximately 0.1 seconds. However,
this really just represents the tests scanning/checking granularity.
The time to finish propagating and committing a single message shows
a slight linear slope upwards from about 0.3 seconds for a three node
cluster to about 0.5 seconds for a 41 node cluster. The time to
propagate 100 messages (Figure~\ref{fig:propagate100}) has a much
higher slope and also appears to be non-linear. TODO: why?

Figure~\ref{fig:cluster_up} shows the results of running the
\emph{test/test\_up.js} test to determine the amount of time that
a cluster takes to start. The cluster is fully started once a
leader has been elected and the leader has successfully added all the
other nodes to the running configuration. Each election timer value
is represented as a different color and the line represents the
average value for each cluster size. The cluster startup times show
a fairly linear increase as the cluster size increases. The reason
that the slope varies with election timer is due to the fact that the
election timer also determines the leader heartbeat value (currently
set to one fifth of the election timer value). The leader heartbeat
determines the frequency with which \emph{appendEntries} messages are
sent from the leader to followers and is also used to propagate log
entries.  This means that the rate at which log entries (and thus
cluster configuration updates) can be propagated is proportional to
the election timer value.

Figures \ref{fig:recovery_leader_failure} and
\ref{fig:recovery_half_failure} show the results of running the
\emph{test/test\_kill\_nodes.js} in both modes (kill only leader and
kill just under half the nodes). Each election timer value is
represented as a different color and the solid line represents the
average recovery time for each cluster size. When only the leader is
killed, the recovery times are constant flat regardless of cluster
size. This is because after a new leader is elected, there is only
a single configuration entry to propagate to the rest of the cluster
(to remove the old failed leader). One interesting thing to note is
that with a 1000ms timer, the 3 node cluster is actually the most
inefficient size because the average time until a leader election
starts is inversely proportional to cluster size so the delay until an
election happens is a significant contributor to the three node
cluster with 1000ms timer. With a larger cluster or shorter election
timer, the contribution of election delay is not noticeable. The
recovery from almost half the nodes failing is not flat because the
amount of configuration information that needs to be propagated is
proportional to the size of the cluster.

At larger cluster sizes (35 and greater) the cluster begins to
experience failures. For example, when testing with a 250ms election
timer, once the cluster size reaches 35 nodes, the cluster is
frequently (more than 50\% of the time) unable to recover from the
half cluster shutdown test. At a cluster size or 41 and with a 1000ms
election timer, the cluster begins to experience difficulty
successfully starting up (about 50\% failure rate). The basic problem
appears to be due to the average election timeout period being too low
in relation to communication latency. During leader election, the
first candidate is unable to gather enough votes to become leader
before other candidates arise and steal votes which prevents a quorum.

Some (or most) of the communication latency may be a result of CPU or
memory contention due to all the browser instances running on a single
host. The use of multiple hosts may show that the system can support
much larger clusters for a given election timeout value. However, with
a fixed election timer value, there is always a cluster size for which
communication latency will begin to prevent the cluster from being
able to perform leader elections. A dynamically adjusted election
timer is something that should be explored especially for a browser
context where the communication latency may be highly variable over
time.
